Defining properties regulating pioneer-factor binding and function

### 2023-02-20 manuscript/manuscript_sections/_04_Discussion.qmd
update manuscript
update manuscript draft

---
format: html
editor: visual
bibliography: ../references.bib
#csl: nature.csl
---

# Discussion

@@

Despite the ability of PFs to bind to closed chromatin, most PFs occupy only a subset of target motifs in the genome and PF binding and activity is often cell-type specific [@Larson2021].
In many cases, the presence of active or repressive chromatin is an important determinant of transcription factor binding, but the extent to which this informs PF binding is unclear.
For the reprogramming factors Oct4, Sox2 and Klf4, and c-Myc (OSKM), high levels of the repressive histone modification H3K9me3 mark regions that are resistant to OSKM binding.
Depletion of H3K9 methyltransferases increases OSKM occupancy at these regions [@Soufi2012].
In the case of Pax7, a PF important for specification of pituitary melanotropes, H3K27me3 is correlated with cell-type specific Pax7 binding [@Mayran2019].
Zld, Grh and Twi binding showed similar correlations with local chromatin structure.
Even when expressed at low levels, all three proteins bind to active, accessible regions with degenerate motifs or no detectable motifs at all.
At approximately physiological levels, these factors bind to canonical motifs in naïve chromatin, and are largely excluded from repressed chromatin (bearing H3K9me3 or H3K27me3).
When we depleted H3K27me3 with the polycomb inhibitor tazemetostat, we observed no effect on binding within these repressed regions.
In contrast to taz treatment, expression of Zld and Grh, but not Twi, at higher levels allowed at least some binding to these repressed regions.
These results suggest that despite the correlation with H3K27me3, protein concentration and affinity for target motifs may be more important than the presence of polycomb silencing for tissue-specific binding and emphasize the importance of functionally testing the importance of different chromatin marks to PF binding.
Cooperative binding with cell-type specific cofactors can also stabilize PF binding to tissue-specific binding sites [@Donaghey2018].
This may also be important for Zld, Grh, and Twi, and may explain the presence of class VI sites that are resistant to binding and lack any repressive histone modifications.
Further exploration of how chromatin structure and cofactor expression influence PF binding and activity will be essential to developing a more complete understanding of how these factors function in development and disease.

>>>>

---
format: html
editor: visual
#bibliography: ../references.bib
---

@@

Despite the ability of PFs to bind to closed chromatin, most PFs occupy only a subset of target motifs in the genome and PF binding and activity is often cell-type specific [@Larson2021].
In many cases, the presence of active or repressive chromatin is an important determinant of transcription factor binding, but the extent to which this informs PF binding is unclear.
For the reprogramming factors Oct4, Sox2 and Klf4, and c-Myc (OSKM), high levels of the repressive histone modification H3K9me3 mark regions that are resistant to OSKM binding.
Depletion of H3K9 methyltransferases increases OSKM occupancy at these regions [@Soufi2012].
In the case of Pax7, H3K27me3 is correlated with cell-type specific Pax7 binding [@Mayran2019].
Zld, Grh and Twi binding showed similar correlations with local chromatin structure.
Even when expressed at low levels, all three proteins bind to active, accessible regions with degenerate motifs or no detectable motifs at all.
At approximately physiological levels, these factors bind to canonical motifs in naïve chromatin, and are largely excluded from repressed chromatin (bearing H3K9me3 or H3K27me3).
When we depleted H3K27me3 with the polycomb inhibitor tazemetostat, we observed no effect on binding within these repressed regions.
In contrast to taz treatment, expression of Zld and Grh, but not Twi, at higher levels allowed at least some binding to these repressed regions.
These results suggest that despite the correlation with H3K27me3, protein concentration and affinity for target motifs may be more important than the presence of polycomb silencing for tissue-specific binding and emphasize the importance of functionally testing the importance of different chromatin marks to PF binding.
Cooperative binding with cell-type specific cofactors can also stabilize PF binding to tissue-specific binding sites [@Donaghey2018].
This may also be important for Zld, Grh, and Twi, and may explain the presence of class VI sites that are resistant to binding and lack any repressive histone modifications.
Further exploration of how chromatin structure and cofactor expression influence PF binding and activity will be essential to developing a more complete understanding of how these factors function in development and disease.

### 2023-02-20 manuscript/manuscript_sections/_06_Figure_Legends.qmd
update manuscript draft

**Extended Data Figure 4: Twist binding leads to chromatin opening and transcriptional activation**

-   Western blot of Twi induction

-   Twi volcano and examples

-   ~~TWI RNA-seq volcano and examples3~~

-   ~~Twi RNA-seq GO analysis~~

-   Genome browser examples of specific and nonspecific accessibility changes?

-   motif enrichment for nonspecific changes?

-   \% of TFs bound and activated by Zld/Grh as indicator of nonspecific changes?

-   Should we include any analysis of decreased/downregulated sites for RNA-seq or ATAC?

-   Violin plots to emphasize correlation between ATAC and RNA-seq?

>>>>

**Extended Data Figure 4: Twist binding leads to chromatin opening and transcriptional activation. a,** Immunoblot showing titration of Twi or HA-Twi protein levels in stable cell lines.
